---
title: ""
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE)

# Load analysis results
loadd(trait_data_for_si)
loadd(climate_ancova_results)
loadd(phylosig_binary_traits_strict)
loadd(correlated_evo_test_strict)
loadd(best_fit_div_models)
loadd(func_comm_struc_combined)

```

***New Phytologist*** **Supporting Information**

Article title: Life in the canopy: Phylogenetic filtering and morphological canalization in epiphytic ferns

Authors: Joel H. Nitta, James E. Watkins, Jr., Ravahere Taputuarai, and Charles C. Davis

Article acceptance date: 

The following Supporting Information is available for this article:

**Table S1** Trait values of ferns on Moorea, French Polynesia. Abbreviations: SLA, specific leaf area. For data source, L = lab observation of gametophyte cultured from spore, F = field observation of gametophyte subsequently identified using DNA barcoding, T = morphology inferred based on taxonomy, M = measurement of a field-collected specimen, etc.

```{r}
trait_data_for_si %>%
  write_csv(here::here("si/Table_S1.csv"))
```

**Table S2** Analysis of covariance (ANCOVA) for climate variables in response to growth habit (epiphytic vs. terrestrial) with elevation as a covariate. Abbreviations: d.f., degrees of freedom; MS, mean squares.

```{r climate-ancova}
# Format and write out table as CSV file
climate_ancova_results %>% 
  mutate(
    p.value = signif(p.value, 2) %>% format.pval(., digits = 1),
    meansq = round_t(meansq, 2),
    term = case_when(
      term == "habit" ~ "Growth habit",
      term == "el" ~ "Elevation",
      term == "habit:el" ~ "Growth habit × elevation",
      TRUE ~ term
    )
  ) %>%
  select(`Response variable` = variable, `Source of variation` = term, `d.f.` = df, MS = meansq, P = p.value) %>%
  write_csv(here::here("si/Table_S2.csv"))

```

**Table S3** Phylogenetic signal in binary traits, strict dataset excluding any species whose traits were scored based on taxonomy. Abbreviations: P_rnd, probability of obtaining the observed value of *D* under random trait evolution; P_bm, probability of obtaining the observed value of *D* under simulation of trait evolution by Brownian motion.

```{r phylosig-binary-strict}
# Format and write out table as CSV file
phylosig_binary_traits_strict %>%
  mutate(D = round_t(D, 2)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("habit", "gemmae", "glands", "hairs","morphotype"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Epiphytic growth", "Gemmae", "Glands", "Hairs", "Cordate morphotype"))) %>%
  arrange(trait) %>%
  select(Trait = trait, `Number of presences` = num_present, `Number of absences` = num_absent, D,
         P_rnd = prob_random, P_bm = prob_brownian) %>%
  write_csv(here::here("si/Table_S3.csv"))
```

**Table S4** Pagel's (1994) test of correlated evolution between binary traits, strict dataset excluding any species whose traits were scored based on taxonomy. Each gametophyte trait was coded as a binary trait and tested for correlated evolution with growth habit (epiphytic vs. terrestrial). Lower log likelihood for the dependent model indicates correlation with growth habit. Abbreviations: LL, log likelihood.

```{r corr-evo-strict}
# Format and write out table as CSV file
correlated_evo_test_strict %>%
  mutate(trait = factor(
    trait, 
    levels = c("gemmae", "glands", "hairs","morphotype"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Gemmae", "Glands", "Hairs", "Morphotype"))) %>%
  arrange(trait) %>%
  select(
    Trait = trait,
    `LL (independent model)` = logL_indep,
    `LL (dependent model)` = logL_dep,
    `Likelihood ratio` = likelihood_ratio,
    `P` = pval
  ) %>%
  write_csv(here::here("si/Table_S4.csv"))
```

**Table S5** Best-fitting models explaining community diversity metrics by climate and growth habit. Models within delta AICc 3 of the best-fitting model shown. Abbreviations: FDiv, functional diversity; FEve, functional evenness; FRic, functional richness; MPD, mean phylogenetic distance; MNTD, mean nearest taxon distance; VPD, vapor pressure deficit, SD, standard deviation; AICc, corrected Akaike Information Criterion. Values of MPD and MNTD are the standard effect sizes calculated from comparison of observed values with null distribution (see Methods). Trait names refer to community-weighted mean values (see Methods).

```{r best-fit-mods}
# Format and write out table as CSV file
best_fit_div_models %>%
  mutate(resp_var = factor(
        resp_var, 
        levels = c("ntaxa", "mpd.obs.z", "mntd.obs.z", 
                   "FDiv", "FEve", "FRic",
                   "dissection", "sla", "stipe", "pinna"))) %>%
  mutate(
    resp_var = lvls_revalue(
        resp_var, 
        new_levels = c("Richness", "MPD", "MNTD", 
                       "FDiv", "FEve", "FRic",
                       "Dissection", "SLA", "Stipe length", "Pinna no.")) 
  ) %>%
  arrange(resp_var, delta_AICc) %>%
  mutate(
    modname = modname %>%
      str_replace_all("\\+", " \\+ ") %>%
      str_replace_all("\\.by\\.", " × ") %>%
      str_replace_all("habit", "Growth habit") %>%
      str_replace_all("temp_mean", "Mean temp.") %>%
      str_replace_all("temp_sd", "SD temp.") %>%
      str_replace_all("temp_max", "Max. temp.") %>%
      str_replace_all("vpd_mean", "Mean VPD") %>%
      str_replace_all("vpd_sd", "SD VPD") %>%
      str_replace_all("vpd_min", "Min. VPD")
  ) %>% 
  select(
    `Metric` = resp_var,
    `Model` = modname,
    AICc,
    `R2` = r2,
    delta_AICc
  ) %>%
  write_csv(here::here("si/Table_S5.csv"))
```

**Fig. S1** Correlation plot comparing results of functional community diversity analyses using different sets of traits as input: **a**, standard effect size of mean phylogenetic distance (MPD~func~) **b**, standard effect size of mean nearest taxon distance (MNTD~func~). For details on calculations of MPD~func~ and MNTD~func~, see Methods. Correlations (values in each tile) calculated with Pearson's correlation coefficient using the R package "corrr". All trait sets were scaled to standard deviation of each trait. The following trait sets were used: "all" (all traits), "sporo" (sporophyte traits; frond length and width, stipe length, rhizome width, pinna number, SLA, degree of lamina dissection), "gameto" (gametophyte traits; gametophyte morphotype, presence or absence of glands, hairs, and gemmae), "size" (sporophyte size-related traits; frond length and width, stipe length, rhizome width), "other" (sporophyte traits not related to size; pinna number, SLA, degree of lamina dissection), "reduced" (all traits but excluding frond length and width, which were highly correlated with stipe length and rhizome diameter). For trait sets that included traits with skewed distributions in their raw values (frond length and width, stipe length, and rhizome width), both log-transformed and untransformed sets were included. The log-transformed "reduced" dataset was used for the main results.

```{r fig-s1}
library(corrr)

# Tweak names for easier viewing
func_comm_struc_combined_renamed <-
func_comm_struc_combined %>%
  mutate(dataset = str_remove_all(dataset, "func_comm_struc_test_fern_traits_")) %>%
  mutate(dataset = str_remove_all(dataset, "\\.pool")) %>%
  mutate(dataset = str_replace_all(dataset, "all_reduced", "reduced")) %>%
  mutate(dataset = str_remove_all(dataset, "scaled_")) %>%
  mutate(dataset = str_replace_all(dataset, "log_", "log-trans_"))

# Don't need log-transformed sets that don't include log-transformed traits
func_comm_struc_combined_renamed <- 
  func_comm_struc_combined_renamed %>%
  filter(
    !dataset %in% c("log-trans_gameto", "log-trans_other")
  )

# Convert to wide format, run corrr: MPD
mpd_corr <-
  func_comm_struc_combined_renamed %>%
  select(dataset, site, habit, mpd.obs.z) %>%
  spread(dataset, mpd.obs.z) %>%
  select(-site, -habit) %>%
  correlate %>% 
  rearrange(absolute = FALSE) # arrange so most correlated variables are closest

# Modified from https://github.com/tidymodels/corrr/issues/35
order <- mpd_corr %>% 
  select(-rowname) %>% 
  colnames()

mpd_corr_plot <-
mpd_corr %>% 
  stretch(na.rm = TRUE) %>%
  mutate_at(c("x", "y"), forcats::fct_relevel, ... = order) %>% 
  ggplot(aes(x, y, fill = r)) +
  geom_tile() +
  geom_text(aes(label = as.character(fashion(r))), color = "white", size = 3) +
  scale_fill_scico(palette = "roma", breaks = c(-1, 1)) +
  labs(x = NULL, y = NULL, title = "a") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  )

# Convert to wide format, run corrr: MNTD
mntd_corr <-
  func_comm_struc_combined_renamed %>%
  select(dataset, site, habit, mntd.obs.z) %>%
  spread(dataset, mntd.obs.z) %>%
  select(-site, -habit) %>%
  correlate %>% 
  rearrange(absolute = FALSE) # arrange so most correlated variables are closest

# Modified from https://github.com/tidymodels/corrr/issues/35
# order <- mntd_corr %>% 
#   select(-rowname) %>% 
#   colnames() %>% 
#   rev()

mntd_corr_plot <-
  mntd_corr %>% 
  stretch(na.rm = TRUE) %>%
  mutate_at(c("x", "y"), forcats::fct_relevel, ... = order) %>% 
  ggplot(aes(x, y, fill = r)) +
  geom_tile() +
  geom_text(aes(label = as.character(fashion(r))), color = "white", size = 3) +
  scale_fill_scico(palette = "roma", breaks = c(-1, 1)) +
  labs(x = NULL, y = NULL, title = "b") +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  )

fig_s1 <- mpd_corr_plot + mntd_corr_plot + plot_layout(nrow = 2, ncol = 1)

ggsave(plot = fig_s1, filename = here("si/Fig_S1.pdf"), height = 8, width = 4, units = "in")
```

