---
title: "Constrained, yet diverse: Phylogenetic and morphological filtering in epiphytic ferns"
bibliography: "references.bib"
mainfont: Times New Roman.ttf
fontsize: 12pt
mainfontoptions: 
- BoldFont=Times New Roman Bold.ttf
- ItalicFont=Times New Roman Italic.ttf
- BoldItalicFont=Times New Roman Bold Italic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: none # Use pandoc default, otherwise csl can't be used for reference formatting
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "new-phytologist.csl"]
    includes:
      in_header: newphyt.sty
header-includes: 
  \usepackage{float}
  \makeatletter\renewcommand*{\fps@figure}{H}\makeatother
editor_options: 
  chunk_output_type: console
nocite: | 
  @Vogel1968, @Wollenweber2000
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE)
library(drake)
library(gt)
library(patchwork)
library(knitr)
library(kableExtra)
library(here)
library(magrittr)
library(scales)
library(tidyverse)
library(conflicted)

conflict_prefer("expand", "tidyr")
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")

# Load analysis results
loadd(best_fit_div_models)
loadd(climate_ancova_results)
loadd(climate_plot)
loadd(climate_select)
loadd(combined_comm_div_plots)
loadd(combined_cwm_plots)
loadd(comm)
loadd(correlated_evo_test)
loadd(correlated_evo_test_gemmae)
loadd(daily_mean_climate)
loadd(div_el_model_summaries)
loadd(div_t_test_results)
loadd(div_metrics_all)
loadd(fern_traits)
loadd(importance_heatmap)
loadd(pca_plot)
loadd(pca_results)
loadd(phy)
loadd(phylosig_binary_traits)
loadd(phylosig_cont_traits)
loadd(pic_results)
loadd(species_list)
loadd(traits_with_tree)
```

<!-- FOR CONVERSION TO MS WORD ONLY: remove page numbers -->
\pagenumbering{gobble}

<!-- Set left justification (better conversion to Word than default full justification) -->
\raggedright

<!-- Set indent to about 3 characters throughout -->
\setlength{\parindent}{3ex}

Joel H. Nitta^1,4^, James E. Watkins, Jr.^2^, Ravahere Taputuarai^3^, and Charles C. Davis^1^

^1^Department of Organismic and Evolutionary Biology and Harvard University Herbaria, Harvard University, Cambridge, MA 02138, U.S.A.

^2^Department of Biology, Colgate University, 13 Oak Drive, Hamilton, New York 13346, U.S.A.

^3^Association Te Rau Ati Ati a Tau a Hiti Noa Tu, B.P. 11553, Mahina, Tahiti, French Polynesia

^4^Current affiliation: Department of Botany, National Museum of Natural History, Smithsonian Institute, Washington, D.C., 20013, U.S.A.

\clearpage

## Summary

- The appearance of angiosperm-dominated forests in the Cretaceous and early Cenozoic had a profound effect on terrestrial biota by creating novel ecological niches. The diversity of modern fern lineages is thought to be a response to angiosperm dominance, radiating both in the canopy and on the forest floor. Despite the significance of epiphytic growth for fern evolution, however, few studies have investigated traits related to epiphytism in ferns while accounting for phylogeny.

- We investigated evolutionary patterns in traits related to epiphytic growth at the species and community level using a phylogeny including representatives of multiple epiphytic fern radiations on the island of Moorea, French Polynesia. 

- We observed phylogenetic conservatism in multiple traits related to epiphytic growth in fern sporophytes and gametophytes. Epiphytes had significantly shorter fronds, smaller rhizomes, and a higher frequency of non-cordate gametophytes. Epiphytic communities were clustered phylogenetically but diverse functionally, indicating that the canopy acts as a filter due to its low humidity to limit size while simultaneously enabling diversification into finely divided niches.

- Our results provide an important phylogenetic perspective on the role of functional traits in ferns, and pave the way for more detailed studies examining how niche space is filled within terrestrial and epiphytic habitats.

**Key words:** community ecology, functional traits, gametophyte, Moorea, Society Islands, Tahiti.

\clearpage

## Introduction

The appearance of angiosperm-dominated forests in the late Cretaceous and early Cenozoic had an immense effect on terrestrial ecosystems [@Niklas1983; @Kenrick1997a; @Lloyd2008]. The heterogeneous niche space of the angiosperm canopy comprising stratified light and humidity levels and a range of substrates for growth became rapidly filled by a variety of epiphytic plants, including mosses [@Fiz-Palacios2011], leafy liverworts [@Feldberg2014], orchids [@Givnish2015b], bromeliads [@Givnish2014a], and ferns [@Schneider2004d]. Diversification rates in some of these groups may have further increased during the Paleocene / Eocene thermal maximum, a period of warmer and wetter climates that may have enabled movement into the canopy [@Schuettpelz2009; @Testo2016a]. The transition to epiphytic growth has been particularly important for fern diversity, with ca. 29% of all species exhibiting this growth form relative to ca. 10% across all vascular plants [@Kress1986]. Major transitions to epiphytism occurred at least five times during the evolution of ferns, each followed by diversification during the late Cretaceous or early Cenozoic [@Schuettpelz2009; @WatkinsJr.2012]. Recent studies [@Sundue2015; @Testo2016a; @Lehtonen2017] have failed to find evidence of rapid diversification characteristic of adaptive radiations associated with epiphytic growth in ferns, but have confirmed gradually increasing rates of diversification in Polypodiaceae [@Sundue2015] and generally support the significance of epiphytic growth as a key innovation for extant fern diversity (@Testo2016a; @WatkinsJr.2012; but see @Lehtonen2017).

Despite the ecological opportunities afforded by life in the canopy, including escape from disturbance [@Page2002] and abundant light for photosynthesis [@Benzing2000], epiphytes must contend with greater stress in the form of increased evaporation and extreme oscillations in temperature relative to terrestrial plants [@Zotz2001; @Lowman2012]. Associated with this change, epiphytic ferns have evolved various morphological adaptations that minimize water loss from their leaves, including shorter stipes [@WatkinsJr.2010], thicker leaves [@Kluge2007; @WatkinsJr.2007b], and protection with scales or hairs [@WatkinsJr.2006b]. Some lineages have specialized adaptations to epiphytic growth, such as humus-collection (e.g., *Asplenium*; @Benzing1990, *Drynaria*; @Janssen2005), vegetative desiccation tolerance (DT) (e.g., Hymenophyllaceae; @Shreve1911, *Pleopeltis*; @Pessin1925), CAM photosynthesis (e.g., *Pyrrosia*; @Wong1976), and symbiotic relationships with ants (e.g., *Antrophyum*; @WatkinsJr.2008, *Lecanopteris*; @Gay1993a). Many of these same adaptations are found in epiphytic seed plants [@Benzing1990]. However, the evolution of epiphytism in ferns was more complex given that ferns rely on a free-living gametophyte. The fern gametophyte is anatomically simple, and differs from the sporophyte in anatomy, morphology, and physiology. Such differences create a system of 'dual ecology' where selective pressures acting on one stage are different than those acting on another. For example, DT is prevalent in gametophytes of multiple lineages of epiphytic ferns, but is rare in sporophytes [@WatkinsJr.2007c]. Despite evidence for adaptations to epiphytism in multiple lineages in both sporophytes and gametophytes, no studies to our knowledge have analyzed the evolution of such traits across the fern phylogeny including both life stages and representatives of all five Cretaceous epiphytic radiations *sensu* [@Schuettpelz2009].

Most studies investigating adaptations to epiphytism have utilized a univariate, cross-species approach, analyzing traits individually between terrestrial and epiphytic species [e.g., @WatkinsJr.2007b; @WatkinsJr.2010]. However, given multiple radiations into the epiphytic niche [@Schuettpelz2009], it is likely that traits related to epiphytism carry at least some degree of phylogenetic signal (i.e., the tendency for closely related species to resemble each other). Therefore, a comparative approach incorporating phylogenetic relationships (and if possible, branch lengths) is appropriate [@Felsenstein1985]. However, such approaches have only recently been applied to studies of epiphytism in ferns, and at relatively limited phylogenetic scales [e.g., @Zhang2014b; @Sundue2015]. 

It should be possible to gain greater insight into the selective strength exerted by the epiphytic habitat by utilizing a community-based approach. If the harsh abiotic conditions of the canopy act as a filter, we expect to find lower functional richness at a community level, which should result in decreased phylogenetic diversity under a scenario of phylogenetic trait conservatism [@Webb2002]. However, the relative strength of this environmental filter may depend on overall abiotic conditions; for example, relative humidity (RH) and temperature also vary with elevation [@Cardelus2006]. Elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kessler2011a]. Functional diversity is also correlated with species richness (few species tend to have exactly the same functional traits; @Schleuter2010). Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and species diversity of epiphytic and terrestrial fern communities.

Here, we investigate the evolution and role in community assembly of functional traits related to epiphytic growth in fern gametophytes and sporophytes on Moorea, French Polynesia. Moorea is an isolated tropical oceanic island with a high phylogenetic diversity of ferns distributed over a steep elevational gradient, making it ideal for a community phylogenetic analysis of functional traits. Specifically, we test the following hypotheses related to epiphytic growth in ferns:

1. The epiphytic environment experiences greater extremes in temperature and humidity than the terrestrial environment, thus acting as a filter on the traits of plants that grow there.

2. Epiphytic ferns have converged upon similar morphologies multiple times in both sporophyte and gametophyte phases.

3. Epiphytic fern communities have lower functional and phylogenetic diversity relative to terrestrial communities due to environmental filtering, and the strength of this filtering varies with changes in environment.

## Materials and Methods

### Study Site

Moorea, French Polynesia (\ang{17;30;}S, \ang{149;50;}W) is a small (\SI{135}{\kilo\meter\squared}) tropical oceanic island more than \SI{5000}{\kilo\meter} from the nearest continental landmass. This distance acts as a strong barrier to dispersal [@Carson1995;@Dassler2001], which combined with the young age of the island (\circa \SI{1.5}{\ma} [@Duncan1976]) has led to a relatively small, yet phylogenetically diverse fern flora (\circa \SI{130}{\species}, 8/11 orders \sensu @PteridophytePhylogenyGroupI2016). Ferns of Moorea include representatives of all five of the Cretaceous epiphytic radiations described by @Schuettpelz2009 (\ie, polygrammoids, elaphoglossoids, asplenioids, vittarioids, and filmy ferns) and their terrestrial relatives, making them suitable for a comparative study of traits related to the evolution of epiphytic growth in ferns.

### Community Survey

A fern community survey was carried out on Moorea comprising 17, \SI{10}{\meter} $\times$ \SI{10}{\meter} plots spanning an elevational gradient from \SIrange{20}{2000}{\meter}, with plots placed \circa every \SI{200}{\meter} [@Nitta2017]. Most plots were located along trails leading to the three main peaks on the island, Mt. Rotui (\SI{899}{\meter}), Mt. Mouaputa (\SI{890}{\meter}), and Mt. Tohiea (\SI{1206}{\meter}). Each plot was divided into 25, \SI{2}{\meter} $\times$ \SI{2}{\meter} subplots. Presence/absence of all sporophytes in each subplot was scored and summed to produce an abundance ranking for each species per plot from zero (not observed) to 25 (observed in all subplots). Epiphytic ferns were sampled by hand to \circa \SI{2}{\meter} on tree trunks or confirmed visually from the ground. Species were categorized as epiphytic or terrestrial based on field observations (see Trait Selection and Measurements below). The list of species in each plot was split into epiphytic and terrestrial groups and treated separately during data analysis, hereafter referred to as 'epiphytic' and 'terrestrial' communities. Vouchers were deposited at GH and UC (abbreviations follow Index Herbariorum [@Thiers2019]).

### Environmental Survey

```{r climate-numbers, results="hide"}
# Count total number of days in climate data
total_days <- daily_mean_climate %>% pull(date) %>% n_distinct()

# Check start and end dates
# - start July 7, 2013
assert_that(daily_mean_climate %>% pull(date) %>% min == "2013-07-07")
# - end July 5, 2014
assert_that(daily_mean_climate %>% pull(date) %>% max == "2014-07-05")
 
# Count total number of sites in climate data
total_sites <-
daily_mean_climate %>%
  select(site, habit) %>%
  unique() %>%
  count(habit)
```

Temperature and RH of epiphytic and terrestrial habitats were measured using Hobo Pro v2 dataloggers with the RS3 Solar Radiation Shield (Onset Corporation, USA). A pair of dataloggers was installed for each plot, one mounted at \circa \SI{2}{\meter} on a tree ('epiphytic datalogger'), and one at ground level ('terrestrial datalogger'), except for the Mt. Tohiea \SI{1200}{\meter} plot, which only had an epiphytic datalogger. Temperature and RH were logged once every \SI{30}{\minute} during two survey periods: a preliminary survey from July 18, 2012 to July 7, 2013 at six plots, and a final survey from July 7, 2013 to July 5, 2014 at 15 plots. Two plots were not included in the environmental survey. We used the 2013–-2014 data for the analysis, with one exception: the terrestrial datalogger at the Mt. Tohiea \SI{393}{\meter} plot malfunctioned and was missing data for a substantial part of the final survey period (March 12--July 5, 2014), so we used data for the same dates from the preliminary survey period for this datalogger instead. Three other dataloggers also failed early during the final survey period, but lacked data from the preliminary period, so were excluded from analysis. Other days missing data for any datalogger during the final survey period were excluded from the dataset. The final dataset included `r total_days` days of data for `r total_sites %>% filter(habit == "epiphytic") %>% pull(n)` epiphytic dataloggers and `r total_sites %>% filter(habit == "terrestrial") %>% pull(n)` terrestrial dataloggers. We calculated vapor pressure deficit (VPD) from relative humidity and temperature using the \package{plantecophys} R package [@Duursma2015].

### Trait Selection and Measurements

We selected several traits relevant to epiphytic growth (\tab \@ref(tab:traits-used) and \fig \@ref(fig:traits-used-image)). We preferred morphological traits over physiological traits for our study because data for morphological traits are easier to gather at the scale of our study (\circa \SI{130}{\species} distributed over an elevational gradient from \SIrange{200}{2000}{\meter}). In addition, plant morphology and physiology are closely linked, and physiological traits such as rates of photosynthesis and water use efficiency are known to be correlated with morphological traits such as stipe length in ferns [@WatkinsJr.2010]. Studying the relationships between ecology and morphological traits is useful from a paleobotanical and climate change perspective, as trends in morphological trait variation in the fossil record can be used to gain insight into past climatic conditions [@Chaloner1997; @McElwain2007a; @McElwain2007b; @Royer2008].

Morphological traits of sporophytes were measured from herbarium specimens for most species, or obtained from the literature when plant material was unavailable. Measurements were made directly on plants in the field for a few species with large fronds. A single leaf or rhizome per individual was measured on one to three separately collected individuals per species, and mean values were calculated for each species. Although trait values can vary within fern species with elevation [@Kessler2007a], our study includes a broad phylogenetic sampling, so we assume intraspecific variation is outweighed by interspecies differences. To measure specific leaf area (SLA; \si{\square\meter\per\kilo\gram}), \circa 10--12, \SI{2}{\milli\meter} or \SI{4}{\milli\meter} \diameter punches were taken of laminae between primary veins for a single leaf per individual using biopsy punches, dried at \SI{60}{\celsius} for \SI{24}{\hour}, and weighed. For species with \textless \SI{2}{\milli\meter} between primary veins, leaf fragments excluding primary veins were obtained by dissection, dried at \SI{60}{\celsius} for \SI{24}{\hour}, weighed, and scanned. ImageJ [@Abramoff2004] software was used to measure area per fragment, and SLA calculated in the same way as leaf punches. 

Morphological traits of gametophytes were all qualitative (either categorical or binary; \tab \@ref(tab:traits-used)). Traits were scored by observing gametophytes that were either collected in the field and identified using DNA barcoding [@Nitta2017] or grown from spores of known species in the lab, taken from the literature, or inferred from taxonomic group (i.e., genus) if documented as a stable taxonomic character. For comparative analysis, gametophyte morphotypes were aggregated into two categories: either cordate (\ie, heart-shaped) or non-cordate. Non-cordate morphotypes include elongate (strap or ribbon \sensu @Farrar2008) and filamentous forms (\fig \@ref(fig:traits-used-image)). Gametophytes were cultivated on Bold's media [@Bold1957] supplemented with Nitch's micronutrients [@Nitsch1951].

Growth habit was coded as a binary trait (epiphytic or terrestrial) based on our field observations. Some species do not fall clearly into either category (\eg, hemi-epiphytes; [@Benzing1990; @Dubuisson2003; @Nitta2009]). We used connection to the soil as the criterion for defining binary growth habit; thus, epipetric plants were treated as epiphytic, and hemi-epiphytes and climbing plants were treated as terrestrial. For the purposes of this study, we treated growth habit as a fixed trait and used these categories to define epiphytic and terrestrial communities. Other studies have demonstrated that exceptional epiphytic growth of terrestrial fern species (and vice-versa) is rare [@Cardelus2006; @Kluge2006; @WatkinsJr.2009], and we do not believe that such rare exceptions would affect our analysis.

Trait values and data sources are summarized in Table S1.

### Statistical analysis

We calculated daily maximum, mean, minimum, and standard variation of temperature, RH, and VPD for each site from the raw climate data, then overall means from the daily values. We analyzed correlation between the mean values with Pearson's correlation coefficient, and retained a set with correlation coefficients less than 0.9 (hereafter, "climatic variables"). Variables relating to VPD, which have no meaning if less than zero, and some other variables that had skewed distributions, were square-root transformed prior to fitting models. We tested for differences between epiphytic and terrestrial dataloggers while accounting for elevation with ANCOVA, using growth habit as the categorical variable and elevation as the covariate. We identified the best-fitting linear model for each climatic variable in response to growth habit and elevation by constructing a set of models including the effect of each independent variable alone and their interactions, then selecting the one with the lowest corrected Akaike Information Criterion (AIC~c~) [@Akaike1973]. A single outlier site with unusually high temperature and low humidity (Mt. Rotui, 830 m) site was excluded from ANCOVA and linear models. Mt. Rotui is an isolated peak with greater exposure than the other two mountains where the rest of the plots were established, which are part of a mountain complex on the interior of the island (J. Nitta, pers. obs.).

For comparative analyses, we used the dated phylogenetic tree of [@Nitta2017], which includes all species from Moorea and some from the neighboring island of Tahiti. We trimmed the tree to species occurring only on Moorea (*n* = `r length(species_list)` \si{\species}) using the \function{drop.tip} function of the \package{ape} package [@Paradis2004] in R [@RCoreTeam2015]. We tested for phylogenetic signal in quantitative traits using two metrics that relate the observed trait values to those expected under a model of trait evolution following Brownian motion (BM), Blomberg's *K* [@Blomberg2003] and Pagel's $\lambda$ [@Pagel1999], with the \function{phylosig} function in the \package{phytools} R package [@Revell2012]. $\lambda$ is a scaling parameter that ranges from 0 (traits evolving randomly, as on a star phylogeny) to 1 (traits evolving along branches according to Brownian motion). \K describes the ratio between the amount of observed variance in traits vs. the amount of variance expected under a BM model of trait evolution: when $\K = 1$, traits are evolving according to BM; when $\K > 1$, traits are more conserved than expected under BM; when $\K < 1$, traits have less phylogenetic signal than expected under BM [@Blomberg2003]. We tested for significance in \K by comparing observed values of \K against a null distribution of \num{1000} trees with trait values shuffled randomly across the tips, and in $\lambda$ by conducting a likelihood ratio test comparing the log likelihoods of the observed value of $\lambda$ vs. $\lambda = 0$ (no phylogenetic signal). We tested for phylogenetic signal in qualitative traits using Fritz and Purvis' *D* [@Fritz2010], which is designed to test for signal in binary traits. Values of \D range from 0 under a model of trait evolution by BM, to 1 for random distribution of traits; negative values (greater trait conservation than BM) and values greater than 1 (trait overdispersion) are possible. We analyzed \D using the \function{phylo.d} function in the \package{caper} R package [@Orme2013], which also conducts a significance test by comparing the observed value of \D with distributions of simulated values produced under one of two models: random shuffling of traits on the tree, or simulation of a binary trait under a BM threshold model. To account for possible circularity in measuring the phylogenetic signal of traits that were scored based on taxonomy, we conducted the same analyses on a dataset excluding species that had been scored this way ("strict" dataset).

```{r check-num-traits}
# Check the number of species included in trait PCA
cont_traits = c("sla", "stipe", "length", "width", 
                "rhizome", "pinna", "dissection")

quant_traits_nsp <-
fern_traits %>%
  # Subset to continuous traits
  select(species, habit, cont_traits) %>%
  # Keep only completely sampled species
  filter(complete.cases(.)) %>%
  nrow
```

We detected significant phylogenetic signal in most traits, so we tested for differences between traits of epiphytic vs. terrestrial species using methods that can test for significant differences between traits associated with a binary character while accounting for phylogeny. For quantitative traits, we used the \function{brunch} function in the \package{caper} package [@Orme2013], which calculates phylogenetically independent contrasts in a quantitative character between alternative states of a binary trait. For qualitative (binary) traits, we used Pagel's test of correlated evolution [@Pagel1994] as implemented with the \function{fitPagel} function in the \package{phytools} package [@Revell2012]. This method compares the likelihood between two alternative models for a pair of binary traits. The first model tests the null hypothesis that rates of evolution of the two traits are independent. The second model tests the alternative hypothesis that the rates of evolution of one trait depends on the other. The likelihoods of obtaining the observed data are calculated for each model, then compared using a log likelihood test. A significantly better fit of the dependent model indicates that evolution of the two traits is correlated. The same analyses were also done on the "strict" dataset.

We investigated differences in multivariate trait space between epiphytes and terrestrial species using Principal Components Analysis (PCA). We used only quantitative traits, and included only species that had no missing data for any traits (*n* = `r quant_traits_nsp` \si{\species}). We first log-transformed traits as necessary to reduce skew, then scaled each trait by its standard deviation. We conducted standard and phylogenetically corrected PCA, using the \function{PCA} and \function{phyl.pca} functions in the \package{FactoMineR} [@Le2008] and \package{phytools} [@Revell2012] R packages, respectively.

We used distance-based methods to analyze phylogenetic and functional diversity. Mean phylogenetic distance (MPD) and mean nearest taxon distance (MNTD) describe the mean distance between all species in a community (MPD) or the mean distance between the closest pairs of species in a community (MNTD) [@Webb2000]. These metrics were originally formulated for phylogenetic distances (thus "phylogenetic" in the name), but can be applied in the same way to trait-based distances, enabling a direct comparison between the two [@Sessa2018]. For functional diversity, we used Gower's dissimilarity coeffecient [@Gower1971] because it can use continuous, binary, and qualitative traits, and accepts missing data. We log-transformed traits as needed to account for skew, scaled each trait by its standard deviation, and excluded frond width and length (both frond size traits highly correlated with stipe length) before calculating the distance matrix using the \function{gowdis} function in the R pacakge \package{FD} [@Laliberte2011], with equal weighting for sporophyte and gametophyte traits. To test for the effect of different trait selection schemes, we compared the results of running the same analyses for various sets of traits with and without log-transformation (Supporting Information).

Raw values of MPD and MNTD are statistically related to species richness [@Swenson2014]; to obtain metrics that are uncorrelated with species richness, we calculated their standard effect sizes (SES) by comparing observed values to a distribution of \num{999} randomized communities (equivalent to the Net Relatedness Index and Nearest Taxon Index, respectively, of @Webb2000 with the sign reversed). Selection of the randomization procedure can strongly bias outcomes [@Gotelli2006;@Ulrich2012]. Our null hypothesis is that fern spores are easily capable of dispersal between terrestrial and epiphytic habitats, and across the relatively short distances separating our plots. We therefore constructed null communities using the "independent swap" algorithm of @Gotelli2000 that maintains species frequency and abundance while randomizing species identities. We included all species of ferns from Moorea and the neighboring island of Tahiti with genetic data available (`r Ntip(phy)` \si{\species} total; [@Nitta2017]) for phylogenetic null communities, and all species with trait data available (`r fern_traits %>%pull(species) %>% n_distinct`) for functional null communities. We used the \function{ses.mpd} and \function{ses.mntd} functions in the R package \package{picante} [@Kembel2010] to calculate phylogenetic and functional diversity metrics. Hereafter, "MPD~phy~" and "MNTD~phy", and "MPD~func~" and "MNTD~func~", will be used to refer to the standard effect sizes of MPD and MNTD as measured with phylogenetic and functional trait distances, respectively.

We tested for differences in diversity metrics between epiphytic and terrestrial communities with a paired \tval-test. We simultaneously tested the effects of climate and growth habit on diversity metrics using a full-subsets multiple regression approach with general additive models (GAMs) as implemented in the \package{FSSgamm} R package [@Fisher2018]. For each diversity metric, we built a set of models (Gaussian family, with site as a random effect) taking into account all combinations of explanatory variables and their pairwise interactions, and ranked these by AIC~c~ to find the models that best explain the data. Our dataset included `r nrow(div_metrics_all) - nrow(climate_select)` communities that lacked climate data and could not be included in full-subsets multiple regression analysis. Since we found that all climatic variables are related to elevation (see Results), we also constructed linear models for each diversity metric using the same model selection procedure as in the climate analysis. We checked for spatial autocorrelation in the model residuals with Moran's *I* [@Moran1950] as implemented in the \package{spdep} R package [@Bivand2013].

### Data accessibility

Code to replicate all analyses, figures, and this manuscript are available at https://github.com/joelnitta/canopylife. Raw and processed data are openly available from the Dryad repository at (http://doi.org/[\*\*\* ADD DOI \*\*\*]) [@Nitta2019].

## Results

### Environmental Survey

Mean, maximum, and standard deviation (SD) of temperature and mean, minimum, and SD of vapor pressure deficit (VPD) were retained after excluding variables with correlation coefficients > 0.9. Variables related to temperature changed with elevation but did not differ significantly between epiphytic and terrestrial dataloggers (\tab S2; \fig \ref{fig:climate}). Variables related to VPD differed by growth habit and elevation. Mean and maximum or minimum temperature and VPD all decreased with elevation, whereas SD of temperature increased, but SD of VPD decreased with elevation. Epiphytic dataloggers had higher values than terrestrial ones for variables related to VPD.

### Principal Components Analysis of Trait Data

```{r pca-numbers}
# Extract summary stats for mention in text

# Cumulative amount of variance explained by first two axes, standard PCA
pca_std_cum_var <-
  pca_results$variance %>% 
  filter(analysis_type == "standard", variance_type == "Cumulative Proportion") %>%
  pull(PC2) %>%
  signif(2)

# Cumulative amount of variance explained by first two axes, phylogenetic PCA
pca_phy_cum_var <-
  pca_results$variance %>% 
  filter(analysis_type == "phylogenetic", variance_type == "Cumulative Proportion") %>%
  pull(PC2) %>%
  signif(2)
```

Terrestrial and epiphytic species are partly differentiated in trait space, occupying mostly distinct areas but overlapping in the middle regardless of method used (standard or phylogenetic PCA) (\fig \@ref(fig:pca)(c, d)). The first two PC axes explained `r percent(pca_std_cum_var)` and `r percent(pca_phy_cum_var)` of the variation in the trait data when analyzed using standard and phylogenetic PCA, respectively. In the standard PCA, size-related traits tended to load on PC1, with SLA, pinna, and dissection on PC2. In the phylogenetic PCA, SLA was strongly loaded on PC1, while dissection, pinna, and other size-related traits were mostly loaded on PC2.

### Phylogenetic Signal

```{r physig-numbers}
habit_d <- phylosig_binary_traits %>% filter(trait == "habit") %>% pull(D) %>% round_t(2)
hairs_d <- phylosig_binary_traits %>% filter(trait == "hairs") %>% pull(D) %>% round_t(2)
gemmae_d <- phylosig_binary_traits %>% filter(trait == "gemmae") %>% pull(D) %>% round_t(2)
glands_d <- phylosig_binary_traits %>% filter(trait == "glands") %>% pull(D) %>% round_t(2)
morphotype_d <- phylosig_binary_traits %>% filter(trait == "morphotype") %>% pull(D) %>% round_t(2)
```

Most measured traits showed some degree of phylogenetic signal, but the strength varied across traits (\tabs \@ref(tab:phylosig-cont), \@ref(tab:phylosig-binary), \fig \@ref(fig:tree)), and for quantitative traits, different results were obtained for $\lambda$ and K. When measured with $\lambda$, most sporophyte traits show phylogenetic signal (as expected under a BM model), with values of $\lambda$ close to 1; only number of pinna pairs had $\lambda$ close to zero (\tab \@ref(tab:phylosig-cont)). However, when measured with \K, only frond width and rhizome diameter showed strong phylogenetic signal (more than expected under BM); other traits had values of $\K < 1$ (\tab \@ref(tab:phylosig-cont)). All binary gametophyte traits showed significant phylogenetic signal, either similar to (hairs, \D = `r hairs_d`) or more conserved (gemmae \D = `r gemmae_d`; glands, \D = `r glands_d`; morphotype, \D = `r morphotype_d`) than expected under a scenario of evolution by BM (\tab \@ref(tab:phylosig-binary)). Growth habit showed phylogenetic signal nearly identical to that expected under BM (\D = `r habit_d`; \tab \@ref(tab:phylosig-binary)). Similar results were obtained using the "strict" dataset (\tab S3).

### Correlation of Traits with Growth Habit

```{r corr-numbers}
stipe_pval <- pic_results %>% filter(trait == "stipe") %>% pull(p.value) %>% signif(2)
width_pval <- pic_results %>% filter(trait == "width") %>% pull(p.value) %>% signif(2)
dissection_pval <- pic_results %>% filter(trait == "dissection") %>% pull(p.value) %>% signif(2)
morphotype_pval <- correlated_evo_test %>% filter(trait == "morphotype") %>% pull(pval) %>% signif(2)
```

Most sporophyte traits, including all those related to plant size, are significantly correlated with growth habit while taking phylogeny into account (\tab \@ref(tab:pics)). Epiphytic species tend to have smaller values for all traits (\tab \@ref(tab:pics); contrasts were made by subtracting epiphytic from terrestrial values). Short stipe length in particular appears to be strongly correlated with epiphytic growth (phylogenetically independent contrasts, \pval = `r stipe_pval`), as well as frond width (\pval = `r width_pval`) and degree of laminae dissection (\pval = `r dissection_pval`). For the gametophyte traits, only morphotype (non-cordate vs. cordate) was significantly correlated with growth habit (Pagel's test of correlated evolution, \pval = `r morphotype_pval`; \tab \@ref(tab:corr-evo)). Similar results were obtained using the "strict" dataset (\tab S4).

### Functional and Phylogenetic Diversity

```{r comm-div-numbers}
# Format model summaries for printing
model_summ_print <-
  div_el_model_summaries %>%
  mutate(r = round_t(r.squared, 2), 
         p = format_pval(p.value))

# Format t-test results for printing
t_test_print <-
  div_t_test_results %>%
  mutate(p = format_pval(p.value))

# Define simple function for extracting value of interest from
# diversity model results
get_mod_result <- function(data, resp_var_select, value) {
  data %>% 
    filter(var == resp_var_select) %>% 
    pull(!!enquo(value))
}

# Make tibble of abundances
abundances <-
  comm %>%
  mutate(abundance = 
           comm %>%
           select_if(is.numeric) %>%
           rowSums()
  ) %>%
  select(species, abundance)

# Join this with trait data
traits_with_abun <- left_join(
  fern_traits, abundances
)

# Get total number of epiphytic and terrestrial species observed in all plots
n_epi <- traits_with_abun %>% filter(abundance > 0, habit == "epiphytic") %>% 
  pull(species) %>% n_distinct
n_ter <- traits_with_abun %>% filter(abundance > 0, habit == "terrestrial") %>% 
  pull(species) %>% n_distinct

# Calculate mean richness by growth habit
mean_rich <-
  div_metrics_all %>%
  group_by(habit) %>%
  summarize(
    mean_rich = mean(ntaxa) %>% round_t(1),
    sd_rich = sd(ntaxa) %>% round_t(1)
  ) %>%
  mutate(stat = glue("{mean_rich} ± {sd_rich}"))
```

In the full-subsets analysis including all climate variables and diversity metrics, growth habit and temperature emerged as the most important variables (\fig \ref{fig:heatmap}). Community-weighted means of frond length, frond width, stipe length, and rhizome diameter were correlated (Pearson's correlation coefficient > 0.9), so of these we only show stipe length. Most of the the best-fitting models included growth habit or the interaction of growth habit and a temperature-related variable (\tab S5). Spatial autocorrelation was not detected for any of the diversity metrics (\tab S5).

Community-weighted mean values (CWMs) were smaller for epiphytic than terrestrial communities for pinna number, and size-related traits (frond length, frond width, stipe length, and rhizome diameter) (\tval-test, all \pval < 0.05; \fig \@ref(fig:cwm-div)). Standard deviation of community-weighted trait values was larger for size-related traits in terrestrial communities, but smaller for other traits such as frond dissection and pinna number. Specific leaf area decreased with elevation, and was lower for epiphytic relative to terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "sla", "r")`, \pval = `r get_mod_result(model_summ_print, "sla", "p")`). Degree of lamina dissection decreased with elevation in epiphytes, but slightly increased in terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "dissection", "r")`, \pval = `r get_mod_result(model_summ_print, "dissection", "p")`). Frond length and width and rhizome diameter decreased slightly with elevation for terrestrial communities, while increasing slightly for epiphytic communities (\fig \@ref(fig:cwm-div)).

We observed a total of `r traits_with_abun %>% filter(abundance > 0) %>% pull(species) %>% n_distinct` species across all plots, including `r n_epi` epiphytic and `r n_ter` terrestrial species. Mean species richness was `r mean_rich %>% filter(habit == "epiphytic") %>% pull(stat)` \si{\species}per plot for epiphytes and `r mean_rich %>% filter(habit == "terrestrial") %>% pull(stat)` \si{\species} per plot for terrestrial species (not significantly different; two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "ntaxa", "p")`; \fig \@ref(fig:comm-div)). Species richness increased  with elevation for both epiphytes and terrestrial communities (\fig \@ref(fig:comm-div)). Epiphytic communities were more diverse than terrestrial ones as measured by both MPD~func~ and MNTD~func~ (two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "mpd.obs.z.func", "p")` and `r get_mod_result(t_test_print, "mntd.obs.z.func", "p")`, respectively). Functional diversity increased increased along the gradient for both terrestrial and epiphytic communities when measured with MPD~func~ (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "mpd.obs.z.func", "r")`, \pval = `r get_mod_result(model_summ_print, "mpd.obs.z.func", "p")`), but not MNTD~func~. Epiphytic communities were more phylogenetically clustered than terrestrial communities overall when measured using MNTD~phy~ (two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "mntd.obs.z.phy", "p")`) but not MPD~phy~ (\pval = `r get_mod_result(t_test_print, "mpd.obs.z.phy", "p")`; \fig \@ref(fig:comm-div)). No significant trend was detected in phylogenetic community structure with elevation in either terrestrial or epiphytic communities (\fig \@ref(fig:comm-div)). 

## Discussion

Here, we show for the first time using a broadly sampled phylogenetic dataset that multiple morphological traits related to epiphytic growth are evolutionarily conserved in ferns. We find that epiphytic ferns are constrained by size, having consistently smaller fronds and rhizomes than terrerstial species. However, contrary to our hypothesis, epiphytic fern communities appear to be more diverse than terrestrial communities. This suggests that environmental filtering plays a strong role in limiting plant size, while allowing for diversification across other trait axes in epiphytes.

### Effect of Phylogeny on Analysis of Epiphytic Traits

Although several other studies have addressed the relationship of functional traits with growth habit in ferns [e.g., @Hietz1998; @Kluge2007; @WatkinsJr.2010], few have accounted for evolutionary history. We find that growth habit shows phylogenetic signal, as do several of the traits associated with it. In one of the few previous studies to analyze phylogenetic signal in functional traits of ferns, @Zhang2014b found that only one of 16 traits showed significant phylogenetic signal. The only trait in common between the current study and that of @Zhang2014b is SLA, which did not show phylogenetic signal in either study. Given the high degree of phylogenetic signal we observe in most traits and that at least some of the traits between our two studies are likely to be correlated, it is surprising that @Zhang2014b did not detect phylogenetic signal in more traits. However, @Zhang2014b only used \K, which tended to indicate less phylogenetic signal than $\lambda$ in our study. In simulation studies, $\lambda$ has lower error (both Type I and Type II) than \K, which can be inconsistent unless the strength of BM is great [@Munkemuller2012]. Our results indicate the importance of including a phylogenetic component, and including multiple measures of phylogenetic signal, when analyzing traits related to epiphytic growth in ferns.

### Sporophyte Traits

Epiphytes were markedly smaller than terrestrial species as measured by stipe and frond length, frond width, and rhizome diameter at both the species and community level (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)). Our results at the local scale are corroborated by @Testo2018a, who also observed abrupt size reduction associated with epiphytic growth in a globally sampled fern phylogeny. Stipe length may be a critical trait controlling whole leaf hydraulic conductance; since ferns have resistive stipes in general, species growing in water-limited habitats are expected to decrease stipe length to minimize resistance [@WatkinsJr.2010]. Our results support previous non-phylogenetically controlled studies that found a prevalence of short stipes in epiphytic ferns [@WatkinsJr.2010; @Creese2011]. However, given that these were all correlated with each other (Pearson's correlation coefficient > 0.9), and similar traits have been shown to scale together previously [@Arcand2008; @Creese2011], careful experimental manipulations are needed to distinguish the role (or lack thereof) of each in epiphytic growth.

```{r sla-numbers, results = "hide"}
# First smallest SLA: Humata_anderssonii
smallest_sla_1 <- fern_traits %>% arrange(sla) %>% slice(1) %>% pull(sla) %>% signif(3)
assert_that(fern_traits %>% arrange(sla) %>% slice(1) %>% pull(species) == "Humata_anderssonii")

# Second smallest SLA: Ctenopterella_blechnoides
smallest_sla_2 <- fern_traits %>% arrange(sla) %>% slice(2) %>% pull(sla) %>% signif(3)
assert_that(fern_traits %>% arrange(sla) %>% slice(2) %>% pull(species) == "Ctenopterella_blechnoides")
```

In contrast to their clearly smaller size, epiphytes were remarkably diverse in other traits. Epiphytes tended to have less dissected fronds with fewer pinna than terrestrial species (\tab \@ref(tab:pics)), but the range of values for these traits was quite wide at the community level (\fig \@ref(fig:cwm-div)). Surprisingly, SLA was only slightly lower in epiphytes relative to terrestrial species (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)). Specific leaf area is an important part of the leaf economic spectrum, which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans [@Wright2004a]. We hypothesized that epiphytes would have lower SLA, given their more extreme growing conditions. However, the parallel relationship in CWM values of SLA with elevation in both epiphytes and terrestrial species (\fig \@ref(fig:cwm-div)) indicates that SLA may be responding more strongly to temperature in both epiphytes and terrestrial species, while only being weakly modified by humidity.

### Gametophyte Traits

Only gametophyte morphotype was found to have a significant association with epiphytic growth. Morphotype was previously observed to correlate with life history and habitat in ferns: terrestrial species tend to have short-lived ($< \SI{1}{\year}$), cordate gametophytes that establish following disturbance and rapidly produce sporophytes, whereas epiphytic species tend to have non-cordate (\eg, elongate or filamentous) gametophytes that persist over multiple growing seasons and produce sporophytes more slowly [@WatkinsJr.2007a; @Farrar2008]. The complex, three-dimensional structure of non-cordate gametophytes slows their drying rates, and may be an adaptation to drought-prone epiphytic habitats [@WatkinsJr.2007c; @Pittermann2013], and promote out-crossing [@Farrar2008]. Our study is the first to our knowledge to demonstrate that the correlation between non-cordate morphology and epiphytism in ferns is significant while controlling for phylogeny. @Sundue2015 found that non-cordate morphology is significantly correlated with increased diversification rates within Polypodiaceae, consistent with the hypothesis that this trait promoted diversification within the epiphytic niche.

We did not find that gemmae production was linked with epiphytism. Gemmae are asexual propagules that allow gametophytes to persist over multiple growing seasons and attain large population sizes [@Farrar2008]. Gemmae production should be a useful trait in an epiphytic context because it may allow gametophytes to increase their surface area and hedge against population loss due to extreme drying events [@Farrar2008]. Gemmae can also function sexually by producing antherida in the presence of antheridiogen secreted by other gametophytes [@Emigh1977], which may confer a selective advantage in the epiphytic environment by enabling outcrossing [@Dassler2001]. It is possible that some, but not all, epiphytic ferns rely on such strategies, and that they are not under similarly strong selection as other traits such as stipe length and gametophyte morphotype. Interestingly, when we classified 'widespread' species as those with gametophytes occurring above or below the minimum or maximum elevational range of conspecific sporophytes using the data of @Nitta2017, we found that gemmae production is correlated with widespread growth (log likelihood dependent model = `r correlated_evo_test_gemmae$dependent.logL %>% round(1) %>% as.numeric`, log likelihood independent model = `r correlated_evo_test_gemmae$independent.logL %>% round(1) %>% as.numeric`, \pval = `r correlated_evo_test_gemmae$P %>% round(3) %>% as.numeric`, Pagel's test of correlated evolution). Thus, it seems that gemmae may function in long-term gametophyte persistence, but are not required for epiphytic growth.

Presence of hairs and glands has strong phylogenetic signal (\tab \@ref(tab:phylosig-binary)), and these traits are correlated with certain taxonomic groups [@Stokey1951; @Nayar1971]. Although hairs have been hypothesized to reduce rates of drying and increase water uptake after desiccation [@Kappen2007], we did not detect a correlation between hair production and epiphytic growth. Similarly, production of glands does not appear to be correlated with epiphytic growth. Wax-secreting glands are prevalent in both sporophytes and gametophytes of many notholaenid ferns, a xeric-adapted clade, but their functional significance has not been established [@Johnson2012]. Further tests are needed to determine if such glands serve a functional role in preventing water loss in desert ferns and epiphytes. 

### Environmental Filtering

Growth habit and temperature are clearly important factors in most of the community diversity metrics we studied, as shown by the full-subsets model analysis (\fig \ref{fig:heatmap}). This is also supported by the linear elevation models (\figs \ref{fig:cwm-div}, \ref{fig:comm-div}), given that temperature shows a strong linear relationship with elevation (\fig \ref{fig:climate}). It is somewhat puzzling that VPD did not contribute significantly to any models, as we hypothesized that this would be important for epiphytes that are frequently subjected to water stress. One reason could be due to conditions at our study site: periods of prolonged rain result in the humidity sensor tip becoming wet, and it does not function normally again (i.e., shows 100% RH) until it dries out. The raw data show several extended periods of 100% RH in terrestrial dataloggers at higher elevations. Despite these possible inaccuracies however, epiphytic habitats are clearly drier than terrestrial (\fig \ref{fig:climate}).

We obtained partial support for our hypothesis that epiphytic fern communities would be more phylogenetically clustered than terrestrial ones; this was observed for MNTD, but not MPD (\fig \@ref(fig:comm-div)). This makes sense given that most epiphytes have radiated within five major clades [@Schuettpelz2009]; each epiphytic community we sampled tends to include closely-related species from two or more epiphytic clades, but has a similar overall level of phylogentic diversity as terrestrial communities. However, we did not detect significant trends in phylogenetic community structure of epiphytic and terrestrial communities along the elevational gradient. It is possible that the scale of the current study (from \SIrange{200}{1200}{\meter}) prevented the detection of such trends. When the plots in this study were analyzed together with additional plots from Tahiti reaching a maximum altitude of just over \SI{2000}{\meter}, @Nitta2017 detected increasing phylogenetic clustering in fern sporophyte communities (epiphytes and terrestrial species together) with elevation. @Kluge2011 also compared phylogenetic and trait diversity between epiphytic and terrestrial ferns over an elevational gradient from \SIrange{100}{3400}{\meter} in Costa Rica, and found trait overdispersion at mid-elevations for epiphytes but no trends in phylogenetic diversity for either group. They accounted this to evolutionary plasticity in traits, but did not quantify the degree of phylogenetic signal in the traits they studied.

Results for functional community diversity contradicted our hypotheses; epiphytic communities were more functionally diverse than terrestrial as measured by both MPD and MNTD (\fig \@ref(fig:comm-div)). However, this result is highly dependent on the traits used to measure functional diversity; when the same analyses were done using only size-related traits (scaled, untransformed stipe length, frond length and width, and rhizome diameter), the opposite result was observed (\fig S1). Futhermore, terrestrial species show a much wider range of values for size traits (\fig \@ref(fig:cwm-div)), and neither functional group appears to be restricted to a smaller area in trait space (\fig \@ref(fig:pca)). This suggests that epiphytic ferns are subject to extreme environmental filtering for size-related traits, while simultaneously diversifying along other trait axes. This could be maintained by niche partitioning within epiphytic habitats. Epiphytic habitats are highly heterogeneous, and microclimatic conditions vary with height along the host and distance from the trunk [@Hietz1995]. The vertical stratification of epiphytic species into different zones along height gradients in host trees is well documented [@Zotz2007]. In a study comparing terrestrial and epiphytic ferns along an elevational gradient from \SIrange{30}{3000}{\meter} in Costa Rica, @WatkinsJr.2006a found that epiphytes tended to sort into low-trunk and canopy zones, with only 18\% found in both. It seems that the canopy habitat limits overall size in epiphytic ferns, but that species with different traits then sort into different niches. The goal of our study was to quantify differences between epiphytic and terrestrial fern communities; future research should focus on bringing comparative phylogenetic and community ecological methods to bear on the large body of work characterizing differences within epiphytic microhabitats.

### Conclusion and Future Directions

We have demonstrated that growth habit, and several traits associated with it, show significant phylogenetic signal in ferns. Furthermore, we find that epiphytes are constrained by size while being remarkably diverse in other traits. Finally, we observed drier conditions in the canopy relative to terrestrial habitats. Taken together, our results support a scenario of environmental filtering acting to limit size and phylogenetic diversity in epiphytic ferns, while simulataneously enabling diversification along other trait axes. This is consistent with a much more broadly sampled study finding that epiphytic and terrestrial ferns do not differ in rates of body size evolution [@Testo2018a]. The generality of this pattern could be further tested by investigating the functional traits of other groups with large epiphytic radiations such as orchids and bromeliads. 

Recent studies [e.g., @Sundue2015; @Testo2016a; @Lehtonen2017; @Testo2018a] have called into question the scenario of rapid Cretaceous radiations of epiphytic ferns posited by @Schuettpelz2009 by failing to detect increased rates of diversification associated with epiphytic growth. Epiphytism itself is likely a complex trait requiring the evolution of multiple associated traits, and the timing and order of the evolution of such traits may be related to the complex diversification dynamics observed in these studies. Small size seems to be a prerequisite for epiphytic growth, as indicated both by our study and @Testo2018a. Here, we identify several traits in addition to plant size (\eg, pinna number, frond dissection, gametophyte morphotype) that may also belong to the suite of epiphytic traits. We hope future studies will build on our results by analyzing traits involved in epiphytic growth other than body size across densely sampled phylogenies to dissect how each may relate to diversification and growth habit. Additionally, future studies should focus on determining in more detail how niche space is partitioned within epiphytic vs. terrestrial habitats, which is clearly a major dividing line for ferns.

## Acknowledgements

We thank members of the Davis Lab, N. Michele Holbrook, and Jonathan Losos for helpful discussion and comments on drafts. Suzanne Vinette, Saad Amer, Tristan Wang, and Tohei Theophilus helped with community surveys and measurement of traits. Staff at the University of California, Berkeley Richard B. Gump South Pacific Research Station, Moorea, French Polynesia, in particular Valentine Brotherson, Hinano Teavai-Murphy, Frank Murphy, and Neil Davies, provided support with obtaining research and collection permits, laboratory space, and logistics. Funding provided in part by National Science Foundation (Doctoral Dissertation Improvement Grant DEB-1311169 to JHN and CCD), Setup Funds from Harvard University to CCD, the American Society of Plant Taxonomists (Research Grant for Graduate Students to JHN), the Garden Club of America (Award in Tropical Botany to JHN), the Harvard University Herbaria (Fernald Fieldwork Fellowship to JHN), the Society of Systematic Biologists (Graduate Student Research Award to JHN), and the Systematics Association (Systematics Research Fund to JHN). 

## Author Contribution

J.H.N. conceived and carried out the study. R.T. collected data. J.H.N. performed the analyses. J.H.N., J.E.W., and C.C.D. wrote the manuscript.

## References
<!-- Insert <div id="refs"></div> to place the refs here instead of at the end -->
<div id="refs"></div>

\clearpage

**\fig \@ref(fig:traits-used-image)**: Examples of traits used in this study. (**a**) Sporophyte traits. All sporophyte traits were quantitative, including frond length and width, stipe length, rhizome dia., number of pinna pairs (in this example, nine) and degree of lamina dissection (in this example, once-pinnate). Specific leaf area (ratio of area to mass of leaf lamina) not depicted. (**b**) Gametophyte traits. Morphotype was defined as a binary trait, either cordate or non-cordate. Examples of non-cordate morphotypes include ribbon (thallus elongate and two-dimensional) and filamentous (thallus single lines of cells). Other gametophyte traits (gemmae, hairs, and glands) were scored as present or absent. Arrows point out instances of each binary trait. Scalebars 1 mm except for glands, which is 0.1 mm. Photographs by J. H. Nitta.

**\fig \@ref(fig:climate)**: Selected microclimatic variables of study plots along an elevational gradient from \SIrange{200}{1200}\meter on Moorea, French Polynesia. Relative humidity (RH) and temperature were recorded every \SI{15}{\minute} with dataloggers from July 7, 2013 to July 5, 2014, and overall means calculated for daily mean, minimum, maximum, and standard deviation (SD). 'Epiphytic' dataloggers were placed at ca. 2 m on trees, and 'terrestrial' dataloggers placed at ground level. Color indicates growth habit: epiphytic dataloggers in green, terrestrial dataloggers in brown. Trendlines fitted using linear models with an interaction between growth habit and elevation (see Methods); all trendlines significant at \pval \textless \num{.05}

**\fig \@ref(fig:pca)**: Principal components analysis (PCA) of traits related to epiphytic growth in ferns from Moorea, French Polynesia. **(a, b)** PC loadings. **(c, d)** species scores. **(a)** and **(c)** are standard PCA, **(b)** and **(d)** are phylogenetic PCA. Epiphytes in green, terrestrial species in brown. Quantitative traits only, including only species with no missing observations for any traits ($\n = 104$ \si{\species}).

**\fig \@ref(fig:tree)**: Time-calibrated phylogenetic tree of ferns from Moorea, French Polynesia with growth habit and associated traits mapped on the tips. Relative value of quantitative (sporophyte) traits shown by heat gradient: low values are lighter, high values are darker. States of qualitative (gametophyte) states and growth habit indicated by colors in key. Grey indicates missing data or non-applicable trait states. Values for three leaf size traits (stipe length, frond length, and frond with) were correlated, so of these, we only present stipe length. Species missing data for six or more traits not shown. For a summary of traits, see \tab \@ref(tab:traits-used). Major Cretaceous epiphytic radiations \sensu @Schuettpelz2009 labeled on tree in green: H, hymenophylloids; T, trichomanoids, V, vittarioids; A, asplenioids; E, elaphoglossoids; P, polygrammoids (trichomanoids diversified prior to the Cretaceous but are included as an important extant epiphytic clade). Units on timescale for tree in millions of years.

**\fig \@ref(fig:cwm-div)**: Change in community-weighted mean (CWM) values of epiphytic and terrestrial fern communities along an elevational gradient on Moorea, French Polynesia. Epiphytic communities in green, terrestrial communities in brown. Error bars are community-weighted standard deviation (values below zero not shown). Trendlines and \rval^2^ shown for significant relationships with elevation or the interaction of elevation and growth habit as determined by a linear model at $\pval<0.05$. \tval shown for significant differences between means of CWMs by growth habit as determined by a two-sided \tval-test at $\pval<0.05$; $*** = \pval<0.001$; $** = \pval<0.01$; $* = \pval<0.05$.

**\fig \@ref(fig:comm-div)**: Functional and phylogenetic diversity of epiphytic and terrestrial fern communities along an elevational gradient on Moorea, French Polynesia. Epiphytic communities in green, terrestrial communities in brown. Response variable abbreviations as follows: MPD~phy~, standard effect size (SES) of mean phylogenetic distance; MNTD~phy~, SES of mean nearest taxon distance; MPD~func~, SES of mean functional distance; MNTD~func~, SES of mean nearest functional distance. Trendlines indicate significant relationships as determined by a linear model at $\pval<0.05$. Boxplots show median values (bold lines) by growth habit. Lower and upper hinges correspond to first and third quartiles, and whiskers extend to values within 1.5 $\times$ the interquartile range. Asterisks indicate statistical significance of two-sided \tval-test; $*** = \pval<0.001$; $** = \pval<0.01$; $* = \pval<0.05$.

\clearpage

```{r traits-used}
# Table 1: traits used in this study

# Read in csv version of the table
traits_used <- read_csv(here("data/table_1.csv"))

# Write out table in rich text format for submitting
# MS as word doc
traits_used %>%
  gt %>%
  fmt_markdown(columns = everything()) %>%
  gt(rowname_col = "trait") %>%
  tab_row_group(
    group = "Sporophyte traits",
    rows = matches("Stipe|Frond|Rhizome|Pinna|Leaf")
  ) %>%
  tab_row_group(
    group = "Gametophyte traits",
    rows = matches("Morphotype|Gemmae|Glands|Hairs")
  ) %>%
  cols_label(
    data_type = "Data type (unit)",
    ecological_significance = "Ecological Significance"
  ) %>% 
  as_rtf() %>%
  write_lines(here("ms/table1.rtf"))

# Include in PDF using kableExtra to convert to latex format
traits_used %>% 
  rename(`Data type (unit)` = data_type, `Functional significance` = ecological_significance) %>%
  column_to_rownames("trait") %>%
  knitr::kable(caption = "Traits used in this study", format = "latex", booktabs = TRUE, escape = TRUE) %>%
  kable_styling(full_width = FALSE) %>%
  pack_rows("Sporophyte traits", 1, 7) %>%
  pack_rows("Gametophyte traits", 8, 11) %>%
  column_spec(3, width = "15em") %>%
  footnote(general = "References: 1, Watkins \\\\textit{et al.} (2010); 2, Vogel (1968); 3, Creese \\\\textit{et al.} (2011); 4, Kluge \\\\& Kessler (2007); 5, Wright \\\\textit{et al.} (2004); 6, Watkins \\\\textit{et al.} (2007b); 7, Pittermann \\\\textit{et al.} (2013); 8, Farrar \\\\textit{et al.} (2008); 9, Wollenweber \\\\& Schneider (2000). Noncordate gametophyte morphotypes include ribbon, strap and filamentous.", general_title = "", threeparttable = TRUE, escape = FALSE)
```

\clearpage

```{r phylosig-cont}
# Table 2: Phylogenetic signal in continuous traits

# Format the raw results for table presentation
phylosig_cont_traits_table <-
phylosig_cont_traits %>%
  mutate_at(vars(kval, lambda), ~round_t(., 2)) %>%
  mutate(lambda_pval = format_pval(lambda_pval)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("stipe", "length", "width", "rhizome", 
               "sla", "pinna", "dissection"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Stipe length", "Frond length", "Frond width", "Rhizome diam.", 
                   "SLA", "Pinna no.", "Frond dissection"))) %>%
  arrange(trait)

# Write out table in rich text format
phylosig_cont_traits_table %>%
  gt(rowname_col = "trait") %>%
  cols_label(
    kval = md("*K*"),
    k_pval = md("*P*(*K*)"),
    lambda_pval = md("*P*(lambda)"),
    lambda = "lambda"
  )  %>% 
  as_rtf() %>%
  write_lines(here("ms/table2.rtf"))

# Include in PDF in latex format
phylosig_cont_traits_table %>%
  rename(K = kval, `P(K)` = k_pval, `P(lambda)` = lambda_pval) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Phylogenetic signal in continuous traits", format = "latex", booktabs = TRUE) %>%
  footnote(general = "K, Blomberg's \\\\textit{K} (Blomberg \\\\textit{et al.}, 2003); lambda, Pagel's lambda (Pagel, 1999). Values of lambda or \\\\textit{K} greater than 1 indicate traits evolving according to Brownian Motion (BM); values of lambda or K near 0 indicate random distribution of traits.", general_title = "", threeparttable = TRUE, escape = FALSE)
```

\clearpage

```{r phylosig-binary}
# Table 3: Phylogenetic signal in binary traits

# Format the raw results for table presentation
phylosig_binary_traits_table <-
phylosig_binary_traits %>%
  mutate(D = round_t(D, 2)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("habit", "gemmae", "glands", "hairs","morphotype"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Epiphytic growth", "Gemmae", "Glands", "Hairs", "Cordate morphotype"))) %>%
  arrange(trait)

# Write out table in rich text format
phylosig_binary_traits_table %>%
  gt(rowname_col = "trait") %>%
  cols_label(
    num_present = "Number of presences",
    num_absent = "Number of absences",
    D = md("*D*"),
    prob_random = md("*P*(rnd)"),
    prob_brownian = md("*P*(bm)")
  ) %>% 
  as_rtf() %>%
  write_lines(here("ms/table3.rtf"))

# Include in PDF in latex format
phylosig_binary_traits_table %>%
  select(trait, `Number of presences` = num_present, `Number of absences` = num_absent,
         D, `P(rnd)` = prob_random, `P(bm)` = prob_brownian) %>%
column_to_rownames("trait") %>%
  kable(caption = "Phylogenetic signal in binary traits", format = "latex", booktabs = TRUE) %>%
  footnote(general = "D, Fritz and Purvis’ \\\\textit{D} (Fritz \\\\& Purvis, 2010); P(rnd), probability of random distribution of traits; P(bm), probability of traits evolving under Brownian motion (BM). Values of \\\\textit{D} 0 or less indicate evolution according to BM; values near 1 indicate random trait distribution.", general_title = "", threeparttable = TRUE, escape = FALSE)
```

\clearpage

```{r pics}
# Table 4: phylogenetically independent contrasts

# Format the raw results for table presentation
pic_table <- 
pic_results %>%
  mutate_at(vars(estimate, std.error, statistic), ~round_t(., 2)) %>%
  mutate(p.value = format_pval(p.value)) %>%
  select(trait, n_contrasts, n_pos_con, statistic, p.value) %>%
  mutate(trait = factor(
    trait, 
    levels = c("stipe", "length", "width", "rhizome", 
               "sla", "pinna", "dissection"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Stipe length", "Frond length", "Frond width", "Rhizome diam.", 
                   "SLA", "Pinna no.", "Frond dissection"))) %>%
  arrange(trait)

# Write out table in rich text format
pic_table %>%
gt(rowname_col = "trait") %>%
  cols_label(
    n_contrasts = "Num. constrasts",
    n_pos_con = "Num. pos. constrasts",
    statistic = md("*t*"),
    p.value = md("*P*"),
  ) %>% 
  as_rtf() %>%
  write_lines(here("ms/table4.rtf"))

# Include in PDF in latex format
pic_table %>%
  rename(t = statistic, P = p.value, `Num. constrasts` = n_contrasts, `Num. pos. constrasts` = n_pos_con) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Results of phylogenetically independent contrasts analysis of quantitative (sporophyte) traits related to epiphytic growth", format = "latex", booktabs = TRUE) %>%
  footnote(general = "All trait contrasts were made by subtracting epiphytic values from terrestrial values; thus, a positive contrast indicates greater values for terrestrial clades.", general_title = "", threeparttable = TRUE)
```

\clearpage

```{r corr-evo}
# Table 5: Pagel's test of correlated evolution

# Format the raw results for table presentation
correlated_evo_test_table <-
correlated_evo_test %>%
  mutate(pval = format_pval(pval)) %>%
  mutate_if(is.numeric, ~round_t(., 1)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("gemmae", "glands", "hairs","morphotype"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Gemmae", "Glands", "Hairs", "Cordate morphotype"))) %>%
  arrange(trait)

# Write out table in rich text format
correlated_evo_test_table %>%
gt(rowname_col = "trait") %>%
  cols_label(
    logL_indep = "LogL(independent)",
    logL_dep = "LogL(dependent)",
    likelihood_ratio = "Likelihood ratio",
    pval = md("*P*"),
  ) %>% 
  as_rtf() %>%
  write_lines(here("ms/table5.rtf"))

# Include in PDF in latex format
correlated_evo_test_table %>%
  rename(`LogL(independent)` = logL_indep, `LogL(dependent)` = logL_dep, `Likelihood ratio` = likelihood_ratio, P = pval) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Pagel's (1994) test of correlated evolution between binary traits", format = "latex", booktabs = TRUE) %>%
  footnote(general = "LogL, log-likelihood. Each gametophyte trait was coded as a binary trait and tested for correlated evolution with growth habit (epiphytic vs. terrestrial growth). Traits with a significantly lower log likelihood for the dependent model are significantly correlated with growth habit.", general_title = "", threeparttable = TRUE)
```

\clearpage

```{r traits-used-image, fig.width=7, fig.height=7, fig.cap = "Examples of traits used in this study."}
knitr::include_graphics(here("images/Figure_3_1_traits_lowres.pdf"))
```

\clearpage

```{r climate, fig.height = 4, fig.width = 6, fig.cap = "Microclimatic variables."}
climate_plot
ggsave(plot = climate_plot, filename = here("ms/fig1.tiff"), height = 4, width = 6, units = "in", dpi = "retina")
```

\clearpage

```{r pca, fig.height = 6, fig.width = 6, fig.cap = "Trait PCA."}
pca_plot
ggsave(plot = pca_plot, filename = here("ms/fig2.tiff"), height = 6, width = 6, units = "in", dpi = "retina")
```

\clearpage

```{r tree, fig.height = 8, fig.width = 6, fig.cap = "Traits on tree."}
traits_with_tree
ggsave(plot = traits_with_tree, filename = here("ms/fig3.tiff"), height = 8, width = 6, units = "in", dpi = "retina")
```

\clearpage

```{r heatmap, fig.height = 4, fig.width = 6, fig.cap = "Variable importance scores."}
importance_heatmap
ggsave(plot = importance_heatmap, filename = here("ms/fig4.tiff"), height = 4, width = 6, units = "in", dpi = "retina")
```

\clearpage

```{r cwm-div, fig.height = 4, fig.width = 8, fig.cap = "Community-weighted means."}
combined_cwm_plots
ggsave(plot = combined_cwm_plots, filename = here("ms/fig5.tiff"), height = 4, width = 8, units = "in", dpi = "retina")
```

\clearpage

```{r comm-div, fig.height = 8, fig.width = 4, fig.cap = "Community diversity metrics."}
combined_comm_div_plots
ggsave(plot = combined_comm_div_plots, filename = here("ms/fig6.tiff"), height = 8, width = 4, units = "in", dpi = "retina")
```
