---
title: "Life in the canopy: Phylogenetic filtering and morphological canalization in epiphytic ferns"
bibliography: "references.bib"
mainfont: Roboto-Regular.ttf
fontsize: 12pt
mainfontoptions: 
- BoldFont=Roboto-Bold.ttf
- ItalicFont=Roboto-Italic.ttf
- BoldItalicFont=Roboto-BoldItalic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: none
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "new-phytologist.csl"]
  bookdown::word_document2:
    toc: no
    fig_caption: yes
    pandoc_args: [ "--csl", "new-phytologist.csl"]
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE)
library(tidyverse)
library(drake)
library(patchwork)
```

Joel H. Nitta^1,4^, James E. Watkins, Jr.^2^, Ravahere Taputuarai^3^, and Charles C. Davis^1^

^1^Department of Organismic and Evolutionary Biology and Harvard University Herbaria, Harvard University, Cambridge, MA 02138, U.S.A.

^2^Department of Biology, Colgate University, 13 Oak Drive, Hamilton, New York 13346, U.S.A.

^3^Association Te Rau Ati Ati a Tau a Hiti Noa Tu, B.P. 11553, Mahina, Tahiti, French Polynesia

^4^Current affiliation: Department of Botany, National Museum of Natural History, Smithsonian Institute, Washington, D.C., 20013, U.S.A.

## Summary

- The appearance of angiosperm-dominated forests in the Cretaceous and early Cenozoic had a profound effect on terrestrial biota by creating novel ecological niches. The diversity of modern fern lineages is thought to be a response to angiosperm dominance, radiating both in the canopy and on the forest floor. Despite the significance of epiphytic growth for fern evolution, however, few studies have investigated traits related to epiphytism in ferns while accounting for phylogeny.

- We investigated evolutionary patterns in traits related to epiphytic growth at the species and community level using a phylogeny including representatives of multiple epiphytic fern radiations on the island of Moorea, French Polynesia.

- We observed phylogenetic conservatism in multiple traits related to epiphytic growth in fern sporophytes and gametophytes. Epiphytes had significantly shorter stipes, smaller rhizomes, and a higher frequency of non-cordate gametophytes while accounting for phylogeny. Epiphytic communities were clustered both phylogenetically and functionally, consistent with the hypothesis that the canopy acts as a filter due to its low humidity.

**Key words:** community ecology, functional traits, gametophyte, Moorea, Society Islands, Tahiti.Our results provide an important phylogenetic perspective on the role of functional traits in ferns, and pave the way for more detailed studies examining how niche space is filled within terrestrial and epiphytic habitats.

## Introduction

The appearance of angiosperm-dominated forests in the late Cretaceous and early Cenozoic had an immense effect on terrestrial ecosystems [@Niklas1983; @Kenrick1997a; @Lloyd2008]. The heterogeneous niche space of the angiosperm canopy comprising stratified light and humidity levels and a range of substrates for growth became rapidly filled by a variety of epiphytic plants, including mosses [@Fiz-Palacios2011], leafy liverworts [@Feldberg2014], orchids [@Givnish2015b], bromeliads [@Givnish2014a], and ferns [@Schneider2004d]. Diversification rates in some of these groups may have further increased during the Paleocene / Eocene thermal maximum, a period of warmer and wetter climates that may have enabled movement into the canopy [@Schuettpelz2009; @Testo2016a]. The transition to epiphytic growth has been particularly important for fern diversity, with ca. 29% of all species exhibiting this growth form relative to ca. 10% across all vascular plants [@Kress1986]. Major transitions to epiphytism occurred at least five times during the evolution of ferns, each followed by diversification during the late Cretaceous or early Cenozoic [@Schuettpelz2009; @WatkinsJr.2012]. Recent studies [@Sundue2015; Testo2016a; @Lehtonen2017] have failed to find evidence of rapid diversification characteristic of adaptive radiations associated with epiphytic growth in ferns, but have confirmed gradually increasing rates of diversification in Polypodiaceae [@Sundue2015] and generally support the significance of epiphytic growth as a key innovation for extant fern diversity (@Testo2016a; @WatkinsJr.2012; but see @Lehtonen2017).

Despite the ecological opportunities afforded by life in the canopy, including escape from disturbance [@Page2002] and abundant light for photosynthesis [@Benzing2000], epiphytes must contend with greater stress in the form of increased evaporation and extreme oscillations in temperature relative to terrestrial plants [@Zotz2001; Lowman2012]. Associated with this change, epiphytic ferns have evolved various morphological adaptations that minimize water loss from their leaves, including shorter stipes [@WatkinsJr.2010], thicker leaves [@Kluge2007; WatkinsJr.2007b], and protection with scales or hairs [@WatkinsJr.2006b]. Some lineages have specialized adaptations to epiphytic growth, such as humus-collection (e.g., *Asplenium*; @Benzing1990, *Drynaria*; @Janssen2005), vegetative desiccation tolerance (DT) (e.g., Hymenophyllaceae; @Shreve1911, *Pleopeltis*; @Pessin1925), CAM photosynthesis (e.g., *Pyrrosia*; @Wong1976), and symbiotic relationships with ants (e.g., *Antrophyum*; @WatkinsJr.2008, *Lecanopteris*; @Gay1993a). Many of these same adaptations are found in epiphytic seed plants [@Benzing1990]. However, the evolution of epiphytism in ferns was more complex given that ferns rely on a free-living gametophyte. The fern gametophyte is anatomically simple, and differs from the sporophyte in anatomy, morphology, and physiology. Such differences create a system of 'dual ecology' where selective pressures acting on one stage are different than those acting on another. For example, DT is prevalent in gametophytes of multiple lineages of epiphytic ferns, but is rare in sporophytes [@WatkinsJr.2007]. Despite evidence for adaptations to epiphytism in multiple lineages in both sporophytes and gametophytes, no studies to our knowledge have analyzed the evolution of such traits across the fern phylogeny including both life stages and representatives of all five Cretaceous epiphytic radiations *sensu* [@Schuettpelz2009].

Most studies investigating adaptations to epiphytism have utilized a univariate, cross-species approach, analyzing traits individually between terrestrial and epiphytic species [e.g., @WatkinsJr.2007b; @WatkinsJr.2010]. However, given multiple radiations into the epiphytic niche [@Schuettpelz2009], it is likely that traits related to epiphytism carry at least some degree of phylogenetic signal (i.e., the tendency for closely related species to resemble each other). Therefore, a comparative approach incorporating phylogenetic relationships (and if possible, branch lengths) is appropriate [@Felsenstein1985]. However, such approaches have only recently been applied to studies of epiphytism in ferns, and at relatively limited phylogenetic scales [e.g., @Zhang2014b; @Sundue2015]. 

It should be possible to gain greater insight into the selective strength exerted by the epiphytic habitat by utilizing a community-based approach. If the harsh abiotic conditions of the canopy act as a filter, we  expect to find lower functional richness at a community level, which should result in decreased phylogenetic diversity under a scenario of phylogenetic trait conservatism [@Webb2002]. However, the relative strength of this environmental filter may depend on overall abiotic conditions; for example, relative humidity (RH) and temperature also vary with elevation [@Cardelus2006]. Elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kessler2011a]. Functional diversity is also correlated with species richness (few species tend to have exactly the same functional traits; @Schleuter2010). Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and species diversity of epiphytic and terrestrial fern communities.

Here, we investigate the evolution and role in community assembly of functional traits related to epiphytic growth in fern gametophytes and sporophytes on Moorea, French Polynesia. Moorea is an isolated tropical oceanic island with a high phylogenetic diversity of ferns distributed over a steep elevational gradient, making it ideal for a community phylogenetic analysis of functional traits. Specifically, we test the following hypotheses related to epiphytic growth in ferns:

1. The epiphytic environment experiences greater extremes in temperature and humidity than the terrestrial environment, thus acting as a filter on the traits of plants that grow there.

2. Epiphytic ferns have converged upon similar morphologies multiple times in both sporophyte and gametophyte phases.

3. Epiphytic fern communities have lower functional and phylogenetic diversity relative to terrestrial communities due to environmental filtering, and the strength of this filtering varies with changes in environment.

## Materials and Methods

## Results

## Discussion

## Acknowledgements

## Figures

```{r climate-plot, fig.height = 4, fig.width = 6, fig.caption = "climate"}
loadd(climate_plot)
climate_plot
```

```{r pca-plot, fig.height = 6, fig.width = 6, fig.caption = "trait PCA"}
loadd(pca_plot)
pca_plot
```

```{r tree, fig.height = 8, fig.width = 6, fig.caption = "traits on tree"}
loadd(traits_with_tree)
traits_with_tree
```

```{r combined-comm-div-plots, fig.height = 8, fig.width = 6, fig.caption = "community diversity metrics"}
loadd(combined_comm_div_plots)
combined_comm_div_plots
```

```{r combined-cwm-div-plots, fig.height = 8, fig.width = 6, fig.caption = "community-weighted means"}
loadd(combined_cwm_plots)
combined_cwm_plots
```

## Tables

## References